Robert C. Dunnell’s _Systematics in Prehistory_ at 50

### 2022-02-16 analysis/paper/paper.Rmd
update some citations

Evolutionary Archaeology is concerned with the study of cultural evolution through material culture proxies and commonly over extended timescales not readily accessible by other means. While evolutionary thinking *per se* is not new in the discipline, an operational evolutionary archaeology emerged only recently, and on the back of several major paradigm transitions [@prentiss2019handbook\]. The notion of major paradigm shifts in the history of archaeology is, however, characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s do stand out as revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's [-@clarke1968a] formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers working within ecological and evolutionary approaches to past culture change are rediscovering the merits of Clarke's conceptual approach in particular, and Binford's extensive data synthesis [@nicholas2012; @lycett2018a]. The year 2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell's [-@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned entirely with classification in archaeology. Appearing at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its goal of addressing a more fundamental issue that was and still is relevant to all branches of archaeology: systematics.

@@

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary deep history: the Palaeolithic. Rooted in French antiquarianism, the development of Paleolithic systematics, for instance, has been likened to so many 'accidents of history' [@clark2009], but there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [-@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's *Systematics in Prehistory* - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

@@

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [-@Cardillo_Charlin_2018\] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. These works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

The South American research stands out as a distinctly coherent and exclusive topic (Topic 59) in our topic model of titles of works citing *Systematics* (Figure \@ref(fig:sip-topic-model)). A topic model is an unsupervised classification of a collection of documents, in this case titles of works citing *Systematics*, that uses a probabilistic approach to generate mixtures of words that represent themes or topics in the collection [-@Roberts_Stewart_Tingley_others_2014; @Roberts_Stewart_Airoldi_2016]. This finds natural groups of topics similar to how clustering on numeric data finds groups of similar items. We generated a topic model with the widely-used Latent Dirichlet Allocation method, which resulted in an optimal number of `r how_many_topics` topics. The topic model shows the persistence of the core themes of *Systematics* in the citing literature, with the most abundant and central topic (Topic 53) about the measurement of material culture variation, and transmission of that variation. The topic model provides additional insights into the contribution of *Systematics* that are not evident in the word frequency analysis. For example, the archaeology of the central and southern Pacific Ocean stands out, and includes works on Polynesian fish hooks, fabrics and stone adzes. Rapa Nui has its own topic, representing the work of Carl Lipo and colleagues on pottery, monuments and population dynamics of that island and the region. Lipo's work is thoroughly grounded in a concern for rigorous and explicit unit construction and innovative testing of archaeological systematics, noting that 'systematics enables us to move beyond common sense' [@lipoisland2021]. We see a concern for the construction of analytical units in areas as diverse as the Palaeolithic of Africa, ceramics of the Maya and Papua New Guinea, and several regions of the Unites States.

The influence of *Systematics* has extended well beyond Dunnell's own primary study area of the North American Southeast. For example, the Portuguese translation published in 2007 has received 75 citations, and the Spanish translation 30. At face value these numbers seem low, but when we take into consideration that Binford's Portuguese translation of *In Pursuit of the Past* published in 1991 has 31 citations, we can have a better perspective of the influence Dunnell's work. On the other hand, in our data we see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [-@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his foreword to the 2002 edition. The low readability of the book has likely limited the accessibility of its contents for readers whose first language is not English, and made it inherently less attractive as a
pedagogical resource.

@@

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by the New Archaeology on the one hand, while also meeting resistance from deeply entrenched traditional approaches to classification. In addition, the volume also appeared just before personal computers were beginning to have a major impact on archaeological research practice [@clarkQuantificationAmericanArchaeology1982; @aldenderfer1987assessing\]. The rigor advocated by Dunnell would have lent itself smoothly to the sorts of computational approaches now increasingly adopted in the discipline. Yet, the approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by economies of personal prestige (Gabriel de Mortillet, 1821--1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of *Systematics* will only become more relevant.

>>>>

Evolutionary Archaeology is concerned with the study of cultural evolution through material culture proxies and commonly over extended timescales not readily accessible by other means. While evolutionary thinking *per se* is not new in the discipline, an operational evolutionary archaeology emerged only recently, and on the back of several major paradigm transitions [@prentiss2019handbook]. The notion of major paradigm shifts in the history of archaeology is, however, characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s do stand out as revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's [@clarke1968a] formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers working within ecological and evolutionary approaches to past culture change are rediscovering the merits of Clarke's conceptual approach in particular, and Binford's extensive data synthesis [@nicholas2012; @lycett2018a]. The year 2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell's [@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned entirely with classification in archaeology. Appearing at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its goal of addressing a more fundamental issue that was and still is relevant to all branches of archaeology: systematics.

@@

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary deep history: the Palaeolithic. Rooted in French antiquarianism, the development of Paleolithic systematics, for instance, has been likened to so many 'accidents of history' [@clark2009], but there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's *Systematics in Prehistory* - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

@@

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [@Cardillo_Charlin_2018] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. These works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

The South American research stands out as a distinctly coherent and exclusive topic (Topic 59) in our topic model of titles of works citing *Systematics* (Figure \@ref(fig:sip-topic-model)). A topic model is an unsupervised classification of a collection of documents, in this case titles of works citing *Systematics*, that uses a probabilistic approach to generate mixtures of words that represent themes or topics in the collection [@Roberts_Stewart_Tingley_others_2014; @Roberts_Stewart_Airoldi_2016]. This finds natural groups of topics similar to how clustering on numeric data finds groups of similar items. We generated a topic model with the widely-used Latent Dirichlet Allocation method, which resulted in an optimal number of `r how_many_topics` topics. The topic model shows the persistence of the core themes of *Systematics* in the citing literature, with the most abundant and central topic (Topic 53) about the measurement of material culture variation, and transmission of that variation. The topic model provides additional insights into the contribution of *Systematics* that are not evident in the word frequency analysis. For example, the archaeology of the central and southern Pacific Ocean stands out, and includes works on Polynesian fish hooks, fabrics and stone adzes. Rapa Nui has its own topic, representing the work of Carl Lipo and colleagues on pottery, monuments and population dynamics of that island and the region. Lipo's work is thoroughly grounded in a concern for rigorous and explicit unit construction and innovative testing of archaeological systematics, noting that 'systematics enables us to move beyond common sense' [@lipoisland2021]. We see a concern for the construction of analytical units in areas as diverse as the Palaeolithic of Africa, ceramics of the Maya and Papua New Guinea, and several regions of the Unites States.

The influence of *Systematics* has extended well beyond Dunnell's own primary study area of the North American Southeast. For example, the Portuguese translation published in 2007 has received 75 citations, and the Spanish translation 30. At face value these numbers seem low, but when we take into consideration that Binford's Portuguese translation of *In Pursuit of the Past* published in 1991 has 31 citations, we can have a better perspective of the influence Dunnell's work. On the other hand, in our data we see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his foreword to the 2002 edition. The low readability of the book has likely limited the accessibility of its contents for readers whose first language is not English, and made it inherently less attractive as a
pedagogical resource.

@@

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by the New Archaeology on the one hand, while also meeting resistance from deeply entrenched traditional approaches to classification. In addition, the volume also appeared just before personal computers were beginning to have a major impact on archaeological research practice [@clarkQuantificationAmericanArchaeology1982; @aldenderfer1987assessing]. The rigor advocated by Dunnell would have lent itself smoothly to the sorts of computational approaches now increasingly adopted in the discipline. Yet, the approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by economies of personal prestige (Gabriel de Mortillet, 1821--1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of *Systematics* will only become more relevant.
